Cell type assignments for SCPCP000015

### 2025-02-07 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
refine plots based on review

- Annotations obtained by running `SingleR` with tumor cells as a reference output by `aucell-singler-annotation.sh`. 
- Consensus cell type annotations output from the `cell-type-consensus` module. 
- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by `run-aucell-ews-signatures.sh`.

@@

```{r}
stopifnot(
  "sce file does not exist" = file.exists(sce_file),
  "aucell results file does not exist" = file.exists(aucell_results_file), 
  "at least one singler file is missing" = all(file.exists(singler_results_files),
  "at least one consensus file is missing" = all(file.exists(consensus_results_files),
)
```

@@

The first thing we will do is compare cell type annotations obtained by each method, `SingleR` with tumor cells as a reference and consensus annotations.

@@

```{r, fig.height=10}
# first get jaccard index between consensus and singler
jaccard_mtx <- make_jaccard_matrix(
  all_info_df,
  "consensus_annotation",
  "singler_annotation"
)

ComplexHeatmap::Heatmap(
  t(jaccard_mtx),
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "SingleR with tumor reference",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  column_title = "Consensus",
  column_title_gp = grid::gpar(fontsize = 12),
  column_names_side = "bottom",
  column_names_gp = grid::gpar(fontsize = 10),
  column_names_rot = 90,
  ## Legend parameters
  heatmap_legend_param = list(
    title = "Jaccard index",
    direction = "vertical",
    legend_width = unit(1.5, "in")
  ))
```

@@

```{r}
# classify based on combined consensus and singler annotation 
all_info_df <- all_info_df |> 
  dplyr::mutate(combined_annotation = dplyr::if_else(
    consensus_annotation == "Unknown" & singler_annotation == "tumor",
    "tumor",
    consensus_annotation
  ),
  # lump for plotting later 
  combined_lumped = combined_annotation |>
      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq()
  )

@@

- `aynaud-ews-targets` > 0.025 AUC
- `staege` > 0.025 AUC
- `wrenn-nt5e-genes` > 0.1 AUC
- `hallmark_EMT` > 0.05 AUC

@@

```{r}
cutoffs = c(
  "auc_aynaud-ews-targets" = 0.025,
  "auc_staege" = 0.025,
  "auc_wrenn-nt5e-genes" = 0.1,
  "auc_hallmark_EMT" = 0.05
)

@@

I don't think these cutoffs are perfect, but I think they are reasonable. 
Any cells that are to the left of these cutoffs will be classified as tumor.

@@

- `tumor EWS-low`: High expression of `wrenn-nt5e-genes` (> 0.1 AUC) and EMT markers `hallmark_EMT` (> 0.05 AUC)
- `tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.025 AUC) and `staege` (>0.025 AUC)

```{r}
# define "final.final" cell types 
all_info_df <- all_info_df |> 
  dplyr::mutate(
    final_annotation = dplyr::case_when(
      `auc_wrenn-nt5e-genes` > 0.1 & auc_hallmark_EMT > 0.05 ~ "tumor EWS-low",
      `auc_aynaud-ews-targets` > 0.025 & auc_staege > 0.025 ~ "tumor EWS-high",
      .default = consensus_annotation
    ),
    # lump together for easier plotting
    final_lumped = final_annotation |>
      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )

@@

For plotting purposes, we will only show the top 10 cell types with fibroblast being the last cell type.

@@

```{r, fig.height=7}
# filter out low expressed genes
dotplot_df <- gene_summary_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 10)

dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(final_lumped), x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Final cell type annotation"
  )

color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = cell_type)) + 
  geom_tile() + 
  scale_fill_brewer(palette = 'Set1') +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

@@

- Generally tumor markers are present throughout, although they are highest in tumor cells. 
- EWS-high proliferative cells show strong expression of proliferation markers. 
- Tumor EWS-low cells show pretty strong expression of the `EWS-low` markers. 
These markers are also present in fibroblasts, but to a lesser degree. 
Additionally, `TNC` (which has been published to be important in the metastatic phenotype in Ewing cells) is specific to the low cells and not found in the fibroblasts, which makes me more confident that those are indeed tumor cells. 
- Endothelial cells show strong expression of `PECAM1` and `VWF`, while that is not seen in most of the other cells. 
- All immune cell types show `PTPRC` as expected with macrophages also showing `MRC1` and T cells showing expression of `CD3G`

@@

```{r, fig.width=7}
# note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
# we also want to specify the annotation name

@@

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  marker_gene_category = validation_markers_df$cell_type,
  col = list(
    marker_gene_category = category_colors
  )
)

>>>>

- Annotations obtained by running `SingleR` with tumor cells as a reference output by [`aucell-singler-annotation.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/aucell-singler-annotation.sh). 
- Consensus cell type annotations output by running [`assign-consensus-celltypes.sh` in `cell-type-consensus`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/assign-consensus-celltypes.sh). 
- AUC values as calculated by `AUCell` for a set of Ewing sarcoma specific gene sets in MSigDB output by `run-aucell-ews-signatures.sh`.

@@

```{r}
stopifnot(
  "sce file does not exist" = file.exists(sce_file),
  "aucell results file does not exist" = file.exists(aucell_results_file), 
  "at least one singler file is missing" = all(file.exists(singler_results_files)),
  "at least one consensus file is missing" = all(file.exists(consensus_results_files))
)
```

@@

The first thing we will do is compare cell type annotations obtained by each method, `SingleR` with tumor cells as a reference (output by [`aucell-singler-annotation.sh`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-ewings/aucell-singler-annotation.sh)) and consensus annotations (output from [`assign-consensus-celltypes.sh` in `cell-type-consensus`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/assign-consensus-celltypes.sh).

@@

```{r, fig.height=5}
# get cell types that have at least 50 cells
singler_to_keep <- all_info_df |> 
  dplyr::count(singler_annotation) |>
  dplyr::filter(n >= 50) |> 
  dplyr::pull(singler_annotation)

# first get jaccard index between consensus and singler
jaccard_mtx <- all_info_df |> 
  dplyr::filter(singler_annotation %in% singler_to_keep) |> 
  make_jaccard_matrix(
    "consensus_annotation",
    "singler_annotation"
  )

ComplexHeatmap::Heatmap(
  t(jaccard_mtx),
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "SingleR with tumor reference",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 12),
  ## Column parameters
  cluster_columns = TRUE,
  column_title = "Consensus",
  column_title_gp = grid::gpar(fontsize = 12),
  column_names_side = "bottom",
  column_names_gp = grid::gpar(fontsize = 12),
  column_names_rot = 90,
  ## Legend parameters
  heatmap_legend_param = list(
    title = "Jaccard index",
    direction = "vertical",
    legend_width = unit(1.5, "in")
  ))
```

@@

```{r}
# classify based on combined consensus and singler annotation 
all_info_df <- all_info_df |> 
  dplyr::mutate(combined_annotation = dplyr::if_else(
    consensus_annotation == "Unknown" & singler_annotation == "tumor",
    "tumor",
    consensus_annotation
  ),
  # lump for plotting later 
  combined_lumped = combined_annotation |>
    forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
    forcats::fct_infreq() |> 
    forcats::fct_relevel("All remaining cell types", after = Inf)
  )

@@

- `aynaud-ews-targets` > 0.04 AUC
- `staege` > 0.01 AUC
- `wrenn-nt5e-genes` > 0.1 AUC
- `hallmark_EMT` > 0.05 AUC

@@

```{r}
cutoffs = c(
  "auc_aynaud-ews-targets" = 0.04,
  "auc_staege" = 0.01,
  "auc_wrenn-nt5e-genes" = 0.1,
  "auc_hallmark_EMT" = 0.05
)

@@

I don't think these cutoffs are perfect, but I think they are reasonable. 
Any cells that are to the right of these cutoffs will be classified as tumor.

@@

- `tumor EWS-low`: High expression of `wrenn-nt5e-genes` (> 0.1 AUC) and EMT markers `hallmark_EMT` (> 0.05 AUC)
- `tumor EWS-high`: High expression of `aynaud-ews-targets` (>0.04 AUC) and `staege` (>0.01 AUC)

```{r}
# top cell type order 
density_plot_order <- c(
  "tumor EWS-high",
  "tumor EWS-low", 
  "Unknown",
  "smooth muscle cell",
  "endothelial cell",
  "macrophage",
  "fibroblast",
  "muscle cell",
  "mature T cell",
  "All remaining cell types"
)

# define "final.final" cell types 
all_info_df <- all_info_df |> 
  dplyr::mutate(
    final_annotation = dplyr::case_when(
      `auc_wrenn-nt5e-genes` > 0.1 & auc_hallmark_EMT > 0.05 ~ "tumor EWS-low",
      `auc_aynaud-ews-targets` > 0.04 & auc_staege > 0.01 ~ "tumor EWS-high",
      .default = consensus_annotation
    ),
    # lump together for easier plotting
    final_lumped = final_annotation |>
      forcats::fct_lump_n(9, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel(density_plot_order)
  )

@@

# first get tally for how many libraries each cell type is present in 
total_libraries <- all_info_df |> 
  dplyr::select(final_annotation, library_id) |> 
  unique() |> # get rid of duplicate cells
  dplyr::group_by(final_annotation) |> 
  dplyr::summarize(
    total_libraries = length(library_id)) |> 
  dplyr::arrange(desc(total_libraries))

# print out both total cells and total libraries 
all_info_df |> 
  dplyr::count(final_annotation, name = "total_cells") |>
  dplyr::left_join(total_libraries, by = "final_annotation") |> 
  dplyr::arrange(desc(total_cells))
```

For plotting purposes, we will only show the top 10 cell types with muscle cell being the last cell type. 
It looks like all other cell types are only present in 1-3 libraries. 
For any cells where there's only 1 cell we may want to confirm that those cells express the expected targets, but that is outside the scope of this notebook.

@@

```{r, fig.height=7, fig.width=10}
# filter out low expressed genes
dotplot_df <- gene_summary_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 10) |>
  dplyr::arrange(final_lumped) |> 
  dplyr::mutate(y_label = as.factor(glue::glue("{final_lumped} ({total_cells})")))

dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma", limits = c(0,2.5), oob = scales::squish) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Final cell type annotation"
  )

# specify legend order 
color_order <- dotplot_df |> 
  dplyr::pull(cell_type) |> 
  unique()

color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = cell_type)) + 
  geom_tile() + 
  scale_fill_brewer(palette = 'Set1', breaks = color_order) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

@@

- Generally tumor markers are present throughout, although they are highest in tumor cells. 
- EWS-high proliferative cells show strong expression of proliferation markers. 
- Tumor EWS-low cells show pretty strong expression of the `EWS-low` markers. 
These markers are also present in fibroblasts, but to a lesser degree. 
Additionally, `TNC`, which has shown to be secreted by Ewing cells and important in the metastatic phenotype of Ewing sarcoma (https://doi.org/10.1016/j.neo.2019.08.007), is specific to the low cells and not found in the fibroblasts, which makes me more confident that those are indeed tumor cells. 
- Endothelial cells show strong expression of `PECAM1` and `VWF`, while that is not seen in most of the other cells. 
- All immune cell types show `PTPRC` as expected with macrophages also showing `MRC1` and T cells showing expression of `CD3G`

@@

```{r, fig.width=10}
# note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
# we also want to specify the annotation name

@@

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  category = validation_markers_df$cell_type,
  col = list(
    category = category_colors
  ),
  annotation_legend_param = list(
      title = "Marker gene category",
      labels = unique(validation_markers_df$cell_type)
    )
)
